Synthesizing Quotient Lenses

### 2018-03-14 papers/QuotientLenses/main.tex
more tweaks to fig 5
fixed text of 4.3
Edited related work section

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}\; (#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R))(w) &=& w \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f})(w) &=& f(w)\\
\canonize(Q_1 \; ; \; Q_2)(w) &=& \canonize(Q_2)(\canonize(Q_1)(w))\\
\canonize(Q_1 \cdot Q_2)(w_1 \cdot w_2) &=& \canonize(Q_1)(w_1) \cdot \canonize(Q_2)(w_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*)(\epsilon) &=& \epsilon \\
\canonize(Q^*)(w_1 \cdot w_2) &=& \canonize(Q)(w_1)\cdot\canonize(Q^*)(w_2) \\[2ex]
\multicolumn{3}{l}{\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})}\\[.5ex]
\multicolumn{3}{l}{\ =\ \ (\canonize(Q_1)(r_1))(w_1) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot(\canonize(Q_n)(r_n))(w_n)}
\end{array}
\]
\end{center}
\caption{QRE canonizers.  Here we assume that the input $w$ to the canonizers
  has been partitioned in the unique way guaranteed to exist by the lens unambiguity conditions.}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{cl}
 & \canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)}) \\
= & (\canonize(Q_1)(r_1)) \cdot (\canonize(Q)(s_1)) \cdot \ldots \cdot
(\canonize(Q)(s_{n-1}))\cdot
(\canonize(Q_n)(r_n))
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

Boomerang's canonizers allow one to canonize a regular language $R$ to by
mapping it to another regular language $S$ which may not be contained in $S$.
Formally, given sets $C$ and $B$ and an equivalence relation on $B$, Foster et
al defined a {\em canonizer} $q$ from $C$ to $B/{\equiv_B}$ to be a pair of
functions $q.\kw{canonize} : C \longrightarrow B$ and $q.\kw{choose} : B
\longrightarrow C$ such that for every $b \in B$,
$$q.\kw{canonize} \; (q.\kw{choose} \; b) \equiv_B \; b$$
This definition gives allows much more latitude for defining canonizers than
QREs. For example, if $\equiv_B = \kw{Tot}(B)$, the equivalence relation that
relates every element in $B$ to every other element in $B$, then every function
from $C$ to $B$ is a canonizer.

@@

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parantheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are unambiguous\BCP{???}, and this
ambiguity can cause 
unintended updates to be made the source, especially when information is
discared in this process. The typechecking phase of a BiFluX program therefore
includes a {\em type normalization} phase that tries to normalize a source types
using automata reduction techniques, and derive a lens between ambiguous and
unambiguous types. On the view side, this phase only tries to normalize view
types into isomorphic types.

{\em Formlenses}~\cite{rajkumar2014lenses} are a variant of lenses that
perform transformations up to equivalence. Formlenses are a bidirectional
generalization of {\em formlets}~\cite{cooper2008essence}. \bcp{The rest of
  this paragraph is hard to understand.  Can we be a little more
  explicit?}Formlets are a high-level abstraction for building Web forms,
and can be represented in Haskell as functions that take a source of
names\bcp{It was puzzling to me, for a while, what an Int had to do with a
  source of names: spell it out}---concretely, an \kw{Int}---as an argument
and produce a triple consisting of an HTML document, a {\em collect}
function, and a modified namesource:
\begin{center}
\begin{lstlisting}
type Formlet a = Int $\to$ (Html, Env $\to$ a, Int)
\end{lstlisting}
\end{center}
The names of any generated form fields are drawn from the namesource, and the
{\em collect} function looks up precisely these names.
Rajkumar et. al. observe that formlets are a useful
abstraction, but they only address one half of the problem---they make it easy
to construct a function that {\em produces} a value of type \codefont{a}, but they do
not provide a way to describe a function that {\em consumes} a value of type
\codefont{a} and embeds it in a form. A different way to think about this problem is
to observe that forms are often used to present an {\em updateable view} of the
data source. Therefore, Rajkumar et. al. leverage the power of bidirectional
transformations to define formlets that behave like updatable views, yielding
the abstraction of formlenses.

The \kw{Formless} type is given by,\bcp{I notice that several displays now
  have commas before them.  If you imagine them typeset inline instead of
  displayed, would this be good grammar?}
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \codefont{Maybe\; a} parameter that provides an optional initial
value of the form, the namesource being changed to a list of integers, and the
{\em collect} function being allowed to return an optional value.

@@

\iffalse
Well-behaved formlenses satisfy properties that are analogous to the classical
\textsc{GetPut} and \textsc{PutGet} law. If {\em extract} is a function from \kw{Html} to \kw{Env} that
extracts an environment containing the names and values of all the form fields
from an HTML document, then a formlens $f$ satisfies the Acceptability law if
for any $v$ of type $a$, $n$ of type \kw{Int} and environment $e$ of type
\kw{Env} if $f$ applied to \kw{Just v} and $n$ yields an HTML file $h$ and a
{\em collect} function $c$ where \kw{extract h} is contained in $e$, then
collect $e$
\fi

Finally, Cunha~\cite{cunha2010relational} uses relational algebra to
encompass many of the existing approaches to bidirectional programming. For
example, using $\cdot$\saz{Why not use $\circ$ or ``;''} to express relational composition, the \textsc{GetPut} law states
that $\get \cdot \lput \subseteq \equiv_V \cdot \pi_1$, where $\pi_1$ is the
projection onto the first factor. In addition to its generality, this
``point-free'' approach to defining bidirectional transformations has the
advantage of creating equational proofs of lens laws that proceed by folding
and unfolding combinator laws.

Next we turn to related work in program synthesis. Much of the research in
synthesis 
assumes that the synthesizer is provided with a collection of examples.
Optician and Optician$_Q$  differ
in that they require the programmer to supply both examples {\em and}
format descriptions in the form of regular expressions or QREs, though we are
far from the only ones to consider type-based synthesis.
There is a trade-off here.  On the one hand, a user must have some programming
expertise to write regular expression (or QRE) specifications, and it requires
some work. On the other hand, such specifications provide a great deal of
information to the synthesis system, which decreases the number of examples
needed (often to zero), makes the system scale well, and allows it to handle
large, complex formats.  By providing these format specifications, the 
synthesis engine does not have to both infer the format of the data as well as
the transformations on it, obviating the need to infer tricky formats like
those involving nested iterations. Furthermore, by focusing on bidirectional
transformations, we limit the space of synthesized functions to bijective ones,
reducing the search space, and the expressiveness of the search space.

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences.
Perhaps the system most similar to Optician is InSynth~\cite{gvero-pldi-2013}, a
system for synthesizing terms in the simply-typed lambda calculus that addresses
equivalences on types. Instead of trying to directly synthesize terms of the
simply-typed lambda calculus, InSynth synthesizes a well-typed term
in the succinct calculus, a language with types
that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure that we used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, we used a pseudo-canonical form, which captures part
of the equivalence relation over types. To preserve completeness, we pushed
some of the remaining parts of the type equivalence relation into a set of
rewriting rules and other parts into the synthesis algorithm itself.

Morpheus~\cite{morpheus} is another synthesis system that uses two
communicating synthesizers to generate programs.  In both Morpheus and
Optician, one synthesizer provides an
outline for the program, and the other fills in that outline with program
details that satisfy the user's specifications.
This approach works well in large search spaces that require some enumerative
search.
One important way that Optician differs from Morpheus is that in
Morpheus, an outline is a sketch---an
\emph{expression}
containing holes---whereas
an outline in Optician is a pair of regular
expressions, i.e., a
\emph{type}.  Moreover, in order to implement an efficient
search procedure, we had to create both a new type language and a new
term language for lenses.  Once we did so, we proved our new, more
constrained language
designed for synthesis was just as expressive as the original, more
flexible and compositional language designed for human programmers.

>>>>

% FOR Regular Expression names
\newcommand{\re}[1]{\ensuremath{\mathtt{#1}}}
\newcommand{\codefont}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\kw}[1]{\textcolor{dkred}{\ensuremath{\mathsf{#1}}}}
\newcommand{\collapse}[2]{\ensuremath{\kw{collapse} \; #1 \mapsto #2}}
\newcommand{\squash}[3]{\ensuremath{\kw{squash} \; #1 \rightarrow #2\; \kw{using} \; #3}}
\newcommand{\perm}[2]{\ensuremath{\kw{perm}(#1)\; \kw{with}\; #2}}
\newcommand{\normalize}[3]{\ensuremath{\kw{normalize}(#1, #2, #3)}}
\newcommand{\eqrel}[1]{\ensuremath{\equiv_{#1}}}
\newcommand{\sep}{\ensuremath{\ | \ }}
\newcommand{\canonize}{\ensuremath{\kw{canonize}}}
\newcommand{\bibtex}{\textsc{Bib}\TeX{}}
\newcommand{\get}{\ensuremath{\kw{get}}}
\newcommand{\semicolon}{\ensuremath{\; ; \;}}
\newcommand{\lput}{\ensuremath{\kw{put}}}
\newcommand{\create}{\ensuremath{\kw{create}}}
\newcommand{\const}{\ensuremath{\kw{const}}}
\newcommand{\swap}{\ensuremath{\kw{swap}}}
\newcommand{\id}{\ensuremath{\kw{id}}}
\newcommand{\lquot}{\ensuremath{\kw{lquot}}}
\newcommand{\rquot}{\ensuremath{\kw{rquot}}}
\newcommand{\lift}{\ensuremath{\kw{lift}}}

@@

Figure~\ref{fig:canonizers} gives the inductive definitions of the
$\canonize{}$ function for each QRE.
\begin{figure}[t]
\begin{center}
\[
\begin{array}{l@\quad l @\quad l}
\canonize(\id(R))(w) &=& w \\
\canonize(\collapse{R}{s})(w) &=& s\\[1ex]
\canonize(\squash{Q_1}{Q_2}{f})(w) &=&
\begin{cases}
\canonize(Q_1)(f(w)) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{otherwise}
\end{cases}\\[3ex]
\canonize(\normalize{R_1}{R_2}{f})(w) &=& f(w)\\
\canonize(Q_1 \; ; \; Q_2)(w) &=& \canonize(Q_2)(\canonize(Q_1)(w))\\
\canonize(Q_1 \cdot Q_2)(w_1 \cdot w_2) &=& \canonize(Q_1)(w_1) \cdot \canonize(Q_2)(w_2)\\[1ex]
\canonize(Q_1 \sep Q_2)(w) &=&
\begin{cases}
\canonize(Q_1)(w) & \text{if } w \in \mathcal{L}(W(Q_1))\\
\canonize(Q_2)(w) & \text{if } w \in \mathcal{L}(W(Q_2))\\
\end{cases}\\[3ex]
\canonize(Q^*)(\epsilon) &=& \epsilon \\
\canonize(Q^*)(w_1 \cdot w_2) &=& \canonize(Q)(w_1)\cdot\canonize(Q^*)(w_2) \\[2ex]
\multicolumn{3}{l}{\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})}\\[.5ex]
\multicolumn{3}{l}{\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1) \cdot \ldots \cdot
\canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)}
\end{array}
\]
\end{center}
\caption{QRE canonizers.  Here we assume that the input $w$ to the canonizers
  has been partitioned in the unique way guaranteed to exist by the lens unambiguity conditions.}
\label{fig:canonizers}
\end{figure}
The \kw{permutation} combinator gives rise to the most interesting
definition:
\[
\begin{array}{l}
\canonize(\perm{Q_1, \ldots, Q_n}{Q})(r_{\sigma(1)}
\cdot s_1 \cdot \ldots \cdot s_{n-1} \cdot r_{\sigma(n)})(w_{\sigma(1)} \cdot s_1
  \cdot \ldots \cdot s_{n-1} \cdot w_{\sigma(n)})\\[.5ex]
\ =\ \ \canonize(Q_1)(r_1)(w_1) \cdot \canonize(Q)(s_1) \cdot \ldots \cdot
\canonize(Q)(s_{n-1})\cdot\canonize(Q_n)(r_n)(w_n)
\end{array}
\]
\noindent 
which places the strings that match $Q_i$ and $Q$ according to the
canonical permutation (i.e., the identity permutation) before applying
the canonizers $\canonize(Q_i)$ and $\canonize(Q)$, respectively.

@@

Boomerang's canonizers allow one to canonize a regular language $R$ to by
mapping it to another regular language $S$ which may not be contained in $S$.
Formally, given sets $C$ and $B$ and an equivalence relation on $B$, Foster et
al defined a {\em canonizer} $q$ from $C$ to $B/{\equiv_B}$ to be a pair of
functions $q.\kw{canonize} : C \longrightarrow B$ and $q.\kw{choose} : B
\longrightarrow C$ such that for every $b \in B$,
$$q.\kw{canonize} \; (q.\kw{choose} \; b) \equiv_B \; b$$
This definition gives allows much more latitude for defining canonizers than
QREs. For example, if $\equiv_B$ is equal to $\kw{Tot}(B)$, the equivalence
relation that relates every element in $B$ to every other element in $B$, then
every function from $C$ to $B$ is a canonizer.

@@

In BiFluX, a regular expression $R$ is said to be unambiguous if there is only
one way to parse a {\em flat} value of type $R$ to a stuctured value of type
$R$; here a flat value of $R$ is a value of $R$ with all left/right tags,
parantheses and list brackets removed. The bidirectional part of BiFluX involves
transforming source regular expressions to view regular expressions so that
data sequences described by the two types can be matched. This process can
result in intermediate types that are ambiguous\BCP{???}\sam{Before I had this
as unambiguous. Was this the issue? Or does this need more explaining even if
the word unambiguous was the issue}, and this ambiguity can cause unintended
updates to be made the source, especially when information is discarded in this
process.

\iffalse
For example, imagine that a book database has source type \\
$S$ = \lstinline{books[book[title[string], author[string]+]*]}
with view type $V$ = \lstinline{title[string]}. When the source type is
transformed to the view type for matching, only titles are kept, the element
labels are destroyed and all authors are replaced for the empty sequence, thus
leaving the intermediate type $S'$ = \lstinline{(title[string],(),()*)*}. In
BiFluX, $V$ is considered a subtype of $S'$ because the flat values of $V$ are
also flat values of $S'$. Since the intermediate type is $S'$ ambiguous, a view
sequence \lstinline{[title["mybook"]]} would be upcast into an intermediate
value \lstinline{[(title["mybook"],((), [ ]))]} that would in turn lead to an
updated source \lstinline{books[[book[(title["mybook"],(aut, [ ]))]]]}, which is
clearly unsatisfactory as it would discard the entire book database except the
first author \lstinline{aut} of the first book.
\fi
The typechecking phase of a BiFluX program therefore includes a {\em type
normalization} phase that tries to normalize a source types using automata
reduction techniques, and derive a lens between ambiguous and unambiguous
types. On the view side, this phase only tries to normalize view types into
isomorphic types. Pacheco et. al. do not claim that their normalization
procedures are complete in the sense that they can disambiguate any ambiguous
type, but they do show that when these procedures succed the normalized types
are unambiguous.

{\em Formlenses}~\cite{rajkumar2014lenses} are a variant of lenses that
perform transformations up to equivalence. Formlenses are a bidirectional
generalization of {\em formlets}~\cite{cooper2008essence}, which are a
high-level abstraction for building Web forms. Formlets encapsulate several
low-level details including selecting field names for elements and parsing data
from client responses. In their work, Rajkumar et. al. represent formlets as
functions that take a source of names---concretely, an integer $n$ that should
be used as the next name---as an argument and produce a triple consisting of an
HTML document, a {\em collect} function, and a modified namesource:
\begin{center}
\begin{lstlisting}
type Formlet a = Int $\to$ (Html, Env $\to$ a, Int)
\end{lstlisting}
\end{center}
The names of any generated form fields are drawn from the namesource, and the
{\em collect} function looks up precisely these names. For example the
\kw{inputInt} combinator below builds a form that accepts an integer,
\begin{center}
\begin{lstlisting}
inputInt :: Formlet Int
inputInt i = let n = show i in
             ($\kw{inputTag}$ {name = n, value = ""},$\lambda e \longrightarrow \kw{read}$ ($\kw{fromJust}$ ($\kw{lookup}$ n e)),i+1)
\end{lstlisting}
\end{center}
where \kw{inputTag} constructs an input element, \kw{lookup}
retrieves a value from an association list \kw{fromJust} extracts the
encapsulated value from a \kw{Maybe} value, and \kw{read} parses an \kw{Int}
from a \kw{String}.

Rajkumar et. al. observe that formlets are a useful abstraction, but they only
address one half of the problem---they make it easy to construct a function
that {\em produces} a value of type \codefont{a}, but they do not provide a way
to describe a function that {\em consumes} a value of type \codefont{a} and
embeds it in a form. A different way to think about this problem is to observe
that forms are often used to present an {\em updateable view} of the data
source. Therefore, Rajkumar et. al. leverage the power of bidirectional
transformations to define formlets that behave like updatable views, yielding
the abstraction of formlenses.

The \kw{Formless} type is given by
\begin{center}
\begin{lstlisting}
type Formlens a = Maybe a $\to$ [Int] $\to$ (Html,Env $\to$ Maybe a,[Int])
\end{lstlisting}
\end{center}
with the differences between this definition and the standard definition for
formlets being an extra \codefont{Maybe\; a} parameter that provides an
optional initial value of the form, the namesource being changed to a list of
integers, and the {\em collect} function being allowed to return an optional
value.

@@

\iffalse
Well-behaved formlenses satisfy properties that are analogous to the classical
\textsc{GetPut} and \textsc{PutGet} law. If {\em extract} is a function from
\kw{Html} to \kw{Env} that extracts an environment containing the names and
values of all the form fields from an HTML document, then a formlens $f$
satisfies the Acceptability law if for any $v$ of type $a$, $n$ of type
\kw{Int} and environment $e$ of type \kw{Env} if $f$ applied to \kw{Just v} and
$n$ yields an HTML file $h$ and a {\em collect} function $c$ where \kw{extract
h} is contained in $e$, then collect $e$
\fi
Another recent line of work that exploits equivalence relations on data is by
Hilken et. al. in \cite{hilken2016testing}. This work is concerned mainly with
testing, and uses {\em Equivalence Partitioning}, a software testing technique
that divides the input data of a software unit into partitions of equivalent
data from which test cases can be derived. This approach has the advantage that
it can greatly reduce the number of test cases and hence the amount of time
spent testing since only one representative from each class is tested.

More concretely, their approach gives the developer an explicit option to
formulat her understanding of two object models being different. The technical
realization is as follows: The developer specifies ``classifying term'', a
closed Object Constraint Language (OCL) query term that can be evaluated in an
object model and returns a characteristic value. Two object models with the same
characteristic value belong to the same equivalence class. For example under a
configuration requiring at least 2 and at most 4 Person objects, the
classifying term \lstinline{Person.allInstances()->size()} would yield three
object models with 2,3 and 4 Person objects respectively.

Classifying terms are therefore analogous to our QREs or to Boomerang canonizers
in the sense that they define an equivalence relation on data. However, while
QREs and canonizers aim at identifying or characterizing the elements that
should be treated equally by a lens, the focus of the work by Hilken et. al. is
on classes that represent specific patterns of particular relevance to the
modeler who is interested in analyzing the behavior of a transformation.

Finally, Cunha~\cite{cunha2010relational} uses relational algebra to encompass
many of the existing approaches to bidirectional programming. For example,
using $\circ$ to express relational composition, the \textsc{GetPut} law states
that $\get \circ \lput \subseteq \equiv_V \circ \pi_1$, where $\pi_1$ is the
projection onto the first factor. In addition to its generality, this
``point-free'' approach to defining bidirectional transformations has the
advantage of creating equational proofs of lens laws that proceed by folding
and unfolding combinator laws.

Next we turn to related work in program synthesis. Much of the research in
synthesis assumes that the synthesizer is provided with a collection of examples.
Optician and Optician$_Q$ differ in that they require the programmer to supply
both examples {\em and} format descriptions in the form of regular expressions
or QREs, though these two systems are far from the only ones to consider
type-based synthesis. There is a trade-off here.  On the one hand, a user must
have some programming expertise to write regular expression (or QRE)
specifications, and it requires some work. On the other hand, such
specifications provide a great deal of information to the synthesis system,
which decreases the number of examples needed (often to zero), makes the system
scale well, and allows it to handle large, complex formats.  By providing these
format specifications, the synthesis engine does not have to both infer the
format of the data as well as the transformations on it, obviating the need to
infer tricky formats like those involving nested iterations.

There are many other recent results showing how to synthesize functions from
type-based
specifications~\cite{augustsson-2004,osera+:pldi15,feser-pldi-2015,scherer-icfp-2015,frankle+:popl16,armando+:pldi16}.
These systems enumerate programs of their target language, orienting their
search procedures to process only terms that are well-typed.
Optician is distinctive in that it synthesizes terms in a language with many
type equivalences. Perhaps the system most similar to Optician is
InSynth~\cite{gvero-pldi-2013}, a system for synthesizing terms in the
simply-typed lambda calculus that addresses equivalences on types. Instead of
trying to directly synthesize terms of the simply-typed lambda calculus,
InSynth synthesizes a well-typed term in the succinct calculus, a language with
types that are equivalent ``modulo isomorphisms of products and
currying''~\cite{gvero-pldi-2013}. The type structure used in Optician
is significantly more complex.  In particular, because Optician types do not
have full canonical forms, Anders et. al. used a pseudo-canonical form that
captures part of the equivalence relation over types were used. To preserve
completeness, they pushed some of the remaining parts of the type equivalence
relation into a set of rewriting rules and other parts into the synthesis
algorithm itself.

Morpheus~\cite{morpheus} is another synthesis system that uses two
communicating synthesizers to generate programs.  In both Morpheus and
Optician, one synthesizer provides an outline for the program, and the other
fills in that outline with program details that satisfy the user's
specifications. This approach works well in large search spaces that require
some enumerative search. One important way that Optician differs from Morpheus
is that in Morpheus, an outline is a sketch---an \emph{expression}
containing holes---whereas an outline in Optician is a pair of regular
expressions, i.e., a \emph{type}.  Moreover, in order to implement an efficient
search procedure, Anders et. al had to create both a new type language and a new
term language for lenses.  Once they did so, they proved their new, more
constrained language designed for synthesis was just as expressive as the
original, more flexible and compositional language designed for human
programmers.
